Supplementary material

### 2014-08-18 manuscripts/manuscript.Rmd
fix supplement to static reference list

```{r cleanup, include=FALSE, cache=FALSE}
write.bibtex(file="components/references.bib")
unlink("simmap.xml")
unlink("birds.xml")
unlink("geospiza.xml")
unlink("rdf_meta.xml")
```

>>>>

```{r cleanup, include=FALSE, cache=FALSE}
unlink("simmap.xml")
unlink("birds.xml")
unlink("geospiza.xml")
unlink("rdf_meta.xml")
```

### 2014-08-18 manuscripts/manuscript.tex
fix supplement to static reference list

This paper has been written as a dynamic document (Xie 2013). The reader
may paste the code blocks shown into their R terminal to follow along
with the exmples shown.

@@

Previously, this was possible using the \texttt{treebase} package
(Boettiger \& {Temple Lang} 2012) for R, which relied on the NEXUS
parser provided in the \texttt{ape} package. Due to inconsistencies
present in the NEXUS standard (Vos \emph{et al.} 2012), a significant
number of TreeBASE NEXUS files could not be read in this way. Using the
more precisely defined NeXML standard, every valid NeXML document on
TreeBASE can now successfully be parsed to R objects.

@@

At the heart of the NeXML standard is the ability to encode arbitrary
metadata in precisely specified, machine-readable formats that maintain
backwards compatibility with any existing NeXML parser. NeXML uses the
``Resource Description Framework in Annotations'', or RDFa, as described
in Vos et al. (2012). These annotations can be used to add information
to any particular element of the data, (a tree, node, edge, or taxonomic
unit; a character matrix or specific trait, or even another metadata
annotation or the NeXML data file as a whole). \texttt{RNeXML} provides
a range of tools for accessing metadata embedded in a NeXML file.

@@

The most powerful way to take advantage of the machine-readable nature
of the metadata involves the construction of SPARQL queries. This
approach can exploit the reasoning and logical deduction capacity of
formal vocabularies (ontologies). For a proper introduction to SPARQL
queries, see Prud'hommeaux (2014) and Hartig (2012).

@@

Here we illustrate this process using the example of stochastic
character mapping (Huelsenbeck \emph{et al.} 2003). A stochastic
character map is simply an annotation of the branches on a phylogeny,
assigning each section of each branch to a particular ``state''
(typically of a morphological characteristic).

@@

Data archiving is increasingly required by journals in evolutionary
biology and biodiversity (e.g. Rausher et al. (2010)). The burden of
preparing and submitting properly annotated data to archives continues
to be a significant barrier (Tenopir et al. (2011), Stodden (2014)), and
many phylogenetic trees upon which studies are based are inaccessible or
lost to the research community (Drew \emph{et al.} 2013).
\texttt{RNeXML} seeks to lower these barriers by providing functions to
immediately archive NeXML files in scientific repositories.

@@

Huelsenbeck, J.P., Nielsen, R. \& Bollback, J.P. (2003). Stochastic
mapping of morphological characters. \emph{Systematic Biology},
\textbf{52}, 131--158. Retrieved from
\url{http://dx.doi.org/10.1080/10635150390192780}

@@

Rausher, M.D., McPeek, M.A., Moore, A.J., Rieseberg, L. \& Whitlock,
M.C. (2010). DATA aRCHIVING. \emph{Evolution}, \textbf{64}, 603--604.
Retrieved from \url{http://dx.doi.org/10.1111/j.1558-5646.2009.00940.x}

@@

Xie, Y. (2013). \emph{Dynamic documents with R and knitr}. Chapman;
Hall/CRC, Boca Raton, Florida. Retrieved from
\url{http://yihui.name/knitr/}

>>>>

This paper has been written as a dynamic document ({\textbf{???}}). The
reader may paste the code blocks shown into their R terminal to follow
along with the exmples shown.

@@

Previously, this was possible using the \texttt{treebase} package
{[}Boettiger \& {Temple Lang} (2012)``{]} for R, which relied on the
NEXUS parser provided in the \texttt{ape} package. Due to
inconsistencies present in the NEXUS standard (Vos \emph{et al.} 2012),
a significant number of TreeBASE NEXUS files could not be read in this
way. Using the more precisely defined NeXML standard, every valid NeXML
document on TreeBASE can now successfully be parsed to R objects.

@@

At the heart of the NeXML standard is the ability to encode arbitrary
metadata in precisely specified, machine-readable formats that maintain
backwards compatibility with any existing NeXML parser. NeXML uses the
``Resource Description Framework in Annotations'', or RDFa, as described
in Vos et al. (2012) or Prud'hommeaux (2014). These annotations can be
used to add information to any particular element of the data, (a tree,
node, edge, or taxonomic unit; a character matrix or specific trait, or
even another metadata annotation or the NeXML data file as a whole).
\texttt{RNeXML} provides a range of tools for accessing metadata
embedded in a NeXML file.

@@

The most powerful way to take advantage of the machine-readable nature
of the metadata involves the construction of SPARQL queries. This
approach can exploit the reasoning and logical deduction capacity of
formal vocabularies (ontologies). For a proper introduction to SPARQL
queries, see Prud'hommeaux (2014) or Hartig (2012)).

@@

Here we illustrate this process using the example of stochastic
character mapping ({\textbf{???}}). A stochastic character map is simply
an annotation of the branches on a phylogeny, assigning each section of
each branch to a particular ``state'' (typically of a morphological
characteristic).

@@

Data archiving is increasingly required by journals in evolutionary
biology and biodiversity (e.g. ({\textbf{???}})). The burden of
preparing and submitting properly annotated data to archives continues
to be a significant barrier (Tenopir \emph{et al.} 2011; Stodden 2014),
and many phylogenetic trees upon which studies are based are
inaccessible or lost to the research community (Drew \emph{et al.}
2013). \texttt{RNeXML} seeks to lower these barriers by providing
functions to immediately archive NeXML files in scientific repositories.

@@

@@

@@

### 2014-08-18 manuscripts/supplement.Rmd
fix supplement to static reference list

library("knitcitations")
cleanbib()
cite_options(citation_format = "pandoc", check.entries=FALSE)
```

@@

While users should consult a formal introduction to SPARQL for further
background, here we illustrate how SPARQL can be used in combination
with R functions in ways that would be much more tedious to assemble
with only traditional/non-semantic queries.  The SPARQL query language
is provided for the R environment through the `rrdf` package `r citep(citation("rrdf"))`,
so we start by loading that package.  We will also make use of functions
from `phytools` and `RNeXML`.

@@

```{r}
rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
tmp <- tempfile()  # rrdf requires a file name, so we must write the XML out first
saveXML(rdf, tmp) 
graph <- load.rdf(tmp)
```

@@

The `RNeXML` package is built for the NeXML Schema, [version 2009],
using the R software environment and the XML package for parsing and
generating XML `r citep(citation("XML"))`,  provides conversions into
the popular `phylo` format provided by the `ape` package `r citep(citation("ape"))`
`RNeXML` can automatically check taxonomic names and add taxonomic
identifiers when generating NeXML using methods from the `taxize` package `r citep(citation("taxize"))`.
Unique identifiers are provided with the `uuid` package `r citep(citation("uuid"))`,
and internet interfaces by the `httr` package `r citep(citation("httr"))`. The
packages `plyr` and `reshape2``r citep(c(citation("plyr"), citation("reshape2")))` are also imported for internal methods .

The `RNeXML` package performs several levels of validation on NeXML files
written from or read into R: validating the XML, and validating against
the current NeXML schema, and performing additional checks through the 
NeXML online validation tool for elements that cannot be enforced by
schema validation alone.  Validation serves to improve consistency in
data encoding which facilitates the use of NeXML across a wide variety
of platforms and software.  A lack of a consistent standard and validation
mechanism has continually plagued the current NEXUS standard and made 
interoperability difficult `r citep("Vos NeXML 2012")`.

Several mechanisms are in place to evaluate the quality of the `RNeXML`
package functions themselves.  In addition to the examples from the
documentation and the automated package checks provided for the R system,
the `RNeXML` package includes a complete test suite using the `testthat`
package `r citep(citation("testthat"))`. Package development on Github includes continuous 
integration through the `travis` platform, in which the full test suite
is run every time the code is changed.  This not only gives users confidence
that the development copy can be successfully installed, but should assist
other developers seeking to contribute to the package to test that their proposed 
changes have not negatively impacted any of the existing code base.

```{r cleanup, include=FALSE, cache=FALSE}
write.bibtex(file="components/sup-references.bib")
unlink("simmap.xml")
unlink("birds.xml")
unlink("geospiza.xml")
unlink(tmp)
```

>>>>

@@

While users should consult a formal introduction to SPARQL for further
background, here we illustrate how SPARQL can be used in combination
with R functions in ways that would be much more tedious to assemble
with only traditional/non-semantic queries.  The SPARQL query language
is provided for the R environment through the `rrdf` package [@Willighagen_2014],
so we start by loading that package.  We will also make use of functions
from `phytools` and `RNeXML`.

@@

```{r}
rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
tmp <- tempfile()  # so we must write the XML out first
saveXML(rdf, tmp) 
graph <- load.rdf(tmp)
```

@@

The `RNeXML` package is built for the NeXML Schema, [version 2009],
using the R software environment and the XML package for parsing and
generating XML [@Temple_Lang_2013],  provides conversions into
the popular `phylo` format provided by the `ape` package [@Paradis_2004],  
`RNeXML` can automatically check taxonomic names and add taxonomic
identifiers when generating NeXML using methods from the `taxize` package [@Chamberlain_2014].
Unique identifiers are provided with the `uuid` package [@Urbank_2014],
and internet interfaces by the `httr` package [@Wickham_2014]. The
packages `plyr` and `reshape2` [@Wickham_2011; Wickham_2007] are also imported for internal methods.

The `RNeXML` package performs several levels of validation on NeXML files
written from or read into R: validating the XML, and validating against
the current NeXML schema, and performing additional checks through the 
NeXML online validation tool for elements that cannot be enforced by
schema validation alone.  Validation serves to improve consistency in
data encoding which facilitates the use of NeXML across a wide variety
of platforms and software.  A lack of a consistent standard and validation
mechanism has continually plagued the current NEXUS standard and made 
interoperability difficult [@Vos_2012].

Several mechanisms are in place to evaluate the quality of the `RNeXML`
package functions themselves.  In addition to the examples from the
documentation and the automated package checks provided for the R system,
the `RNeXML` package includes a complete test suite using the `testthat`
package [@Wickham_2011a]. Package development on Github includes continuous 
integration through the `travis` platform, in which the full test suite
is run every time the code is changed.  This not only gives users confidence
that the development copy can be successfully installed, but should assist
other developers seeking to contribute to the package to test that their proposed 
changes have not negatively impacted any of the existing code base.

```{r cleanup, include=FALSE, cache=FALSE}
unlink("simmap.xml")
unlink("birds.xml")
unlink("geospiza.xml")
unlink(tmp)
```
